On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-14 joint_inf.tex
pass through the fixed x' and y' part.

Consider only the Farris-tree topology.
We will consider two settings, corresponding to whether some branch lengths are known ahead of time or not.

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
If $x'$ and $y'$ are fixed, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w')
\end{align*}
by collapsing like terms and ignoring terms with only $x'$ and $y'$.
Similarly, the other two possibilities for the likelihood are both proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w').
\end{align*}
Since
$$
\log(1+w') >= (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, we focus on the second likelihood.
Simplifying and suppressing some variables, this likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
\begin{align*}
f(w') &= w^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2), \\
\end{align*}
implying $\ell'$ has two zeros according to the quadratic formula.
By $\alpha_1 > \alpha_2$, $\ell'(0) > 0$ implies $\ell$ is increasing at $w'=0$.
We use this fact to show that, if the smaller of the zeros of $f(w')$ is greater than one, $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

>>>>

We will now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
We will consider two settings on the Farris topology, corresponding to whether some branch lengths are known ahead of time or not.
As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
%EM We will show XXX... and introduce \hat{w'} here.

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
Similarly, the other two possibilities for the likelihood are both proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w').
\end{align*}
Since
$$
\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
%EM Not sure what's being suppressed here. Sounds naughty.
Simplifying and suppressing some variables, this likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log likelihood is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite.
%EM I don't think you mean "any inequality"... what about sin(ell') > 0?
any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
\begin{align*}
f(w') &= (w')^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w'\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2),
\end{align*}
%EM Is this obvious at this point? Couldn't b^2 - 4ac be zero in principle?
implying $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w'=0$.
This implies that if the smaller of the zeros of $f(w')$ is greater than one, then $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.
